Exploring dimensions of biodiversity in Japanese ferns

### 2021-07-26 ms/SI.Rmd
Update MS
add maps of genera, barplot of % apo

# Load packages only used when knitting Rmd
library(kableExtra)
```

@@

```{r load-data-si, cache = FALSE}
tar_load(c(
  "japan_shp",
  "comm_scaled_list",
  "occ_point_data_ferns",
  "biodiv_ferns_spatial",
  "biodiv_ferns_endemic_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "traits_nmds",
  "ppgi",
  "phy_sig_results",
  "binary_sig_results",
  "lrt_results",
  "japan_shp",
  "morans_i",
  "t_test_results",
  "aic_env_repro"
))
```

@@

@@

```{r envir, fig.width = 7, fig.height = 9}

# Restrict sites to only those used in model
# (drops outlier % apomictic site)
biodiv_ferns_for_plotting <-
  biodiv_ferns_spatial %>%
  drop_apo_outlier

# crop high-res map of Japan to spatial div results area
japan <- sf::st_crop(japan_shp, sf::st_bbox(biodiv_ferns_for_plotting))

# Set CRS
japan_crs <- sf::st_crs(japan)
biodiv_ferns_for_plotting <- sf::st_set_crs(biodiv_ferns_for_plotting, japan_crs)

# Define function for plotting environment
plot_env <- function(part, label, var_select, fill_lab = "", labels = ggplot2::waiver()) {
  ggplot() +
    geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
    geom_sf(data = biodiv_ferns_for_plotting, aes(fill = .data[[var_select]]), color = "transparent", alpha = 0.8) +
    annotate(
      geom="richtext", y = Inf, x = -Inf, 
      label = glue("<span style='color:black'>**{part}** {label}</span>"), 
      hjust = "inward", vjust = "inward", color = "white") + 
    labs(fill = fill_lab) +
    scale_fill_scico(palette = "lajolla", labels = labels) +
    scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
    theme_grey(base_size = 10) +
    theme(
      panel.grid.major = element_line(color = "grey60", size = 0.1),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "transparent"),
      axis.ticks = element_blank(),
      axis.text = element_text(color = "grey60"),
      axis.title = element_blank(),
      legend.position = "right",
      legend.title = element_markdown()
    )
}

@@

>>>>

# Load packages only used when knitting Rmd
library(kableExtra)
library(scales)
```

@@

```{r load-data-si, cache = FALSE}
tar_load(c(
  "japan_shp",
  "comm_scaled_list",
  "occ_point_data_ferns",
  "biodiv_ferns_spatial",
  "biodiv_ferns_endemic_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "traits_nmds",
  "ppgi",
  "phy_sig_results",
  "binary_sig_results",
  "lrt_results",
  "japan_shp",
  "morans_i",
  "t_test_results",
  "aic_env_repro",
  "comm_ferns",
  "repro_data"
))
```

@@

# Subset ferns biodiversity dataframe: restrict sites to only those used in model
# (drops outlier % apomictic site)
biodiv_ferns_for_plotting <-
  biodiv_ferns_spatial %>%
  drop_apo_outlier

@@

```{r envir, fig.width = 7, fig.height = 9}
# Define function for plotting environment
plot_env <- function(part, label, var_select, fill_lab = "", labels = ggplot2::waiver()) {
  ggplot() +
    geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
    geom_sf(data = biodiv_ferns_for_plotting, aes(fill = .data[[var_select]]), color = "transparent", alpha = 0.8) +
    annotate(
      geom="richtext", y = Inf, x = -Inf, 
      label = glue("<span style='color:black'>**{part}** {label}</span>"), 
      hjust = "inward", vjust = "inward", color = "white") + 
    labs(fill = fill_lab) +
    scale_fill_scico(palette = "lajolla", labels = labels) +
    scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
    theme_grey(base_size = 10) +
    theme(
      panel.grid.major = element_line(color = "grey60", size = 0.1),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "transparent"),
      axis.ticks = element_blank(),
      axis.text = element_text(color = "grey60"),
      axis.title = element_blank(),
      legend.position = "right",
      legend.title = element_markdown()
    )
}

@@

```{r genus-map, fig.width = 7, fig.height = 9}
# Define function for plotting genus-level abundance.
# assumes `comm_ferns`, `japan` are in environment
plot_genus <- function(part, genus_select) {

  genus_abun <-
    comm_ferns %>%
    rownames_to_column("grids") %>%
    as_tibble() %>%
    pivot_longer(-grids, names_to = "species", values_to = "abun") %>%
    filter(str_detect(species, genus_select)) %>%
    group_by(grids) %>%
    summarize(richness = sum(abun), .groups = "drop") %>%
    left_join(select(biodiv_ferns_spatial, -richness), ., by = "grids")

  ggplot() +
    geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
    geom_sf(data = genus_abun, aes(fill = richness), color = "transparent", alpha = 0.8) +
    annotate(
      geom="richtext", y = Inf, x = -Inf, 
      label = glue("<span style='color:black'>**{part}** *{genus_select}*</span>"), 
      hjust = "inward", vjust = "inward", color = "white") + 
    labs(fill = "Richness") +
    scale_fill_scico(palette = "lajolla") +
    scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
    theme_grey(base_size = 10) +
    theme(
      panel.grid.major = element_line(color = "grey60", size = 0.1),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "transparent"),
      axis.ticks = element_blank(),
      axis.text = element_text(color = "grey60"),
      axis.title = element_blank(),
      legend.position = "right",
      legend.title = element_markdown()
    )
}

a <- plot_genus("A", "Equisetum")
b <- plot_genus("B", "Deparia")
c <- plot_genus("C", "Cyrtomium")
d <- plot_genus("D", "Dryopteris")
e <- plot_genus("E", "Pteris")
f <- plot_genus("F", "Diplazium")

**`r s_figure("genus-map")`**. Map of species richness for selected genera of ferns of Japan. **A**, *Equisetum* is an example of a possible refugial lineage with greater species richness in northern Japan. **B**, *Deparia* is an example of a genus with possible recent radiation in southern Honshu. **C**--**F** show genera with high rates of apomictic taxa (`r s_figure("apo-genus")`), which tend to be greatest in southern Honshu.

```{r apo-genus, fig.width = 4, fig.height = 3}
# Bar plot of % apo taxa by genus
percent_apo_genus <-
repro_data %>%
  # Remove NA values, non-ferns
  filter(!is.na(apomict)) %>%
  filter(str_detect(taxon, "Lycopodiella|Lycopodium|Isoetes", negate = TRUE)) %>%
  # Summarize percent apo by genus
  mutate(genus = str_split(taxon, "_") %>% map_chr(1)) %>%
  group_by(genus) %>%
  add_count() %>%
  group_by(genus, n) %>%
  summarize(total_apo = sum(apomict), .groups = "drop") %>%
  mutate(percent_apo = total_apo / n) %>%
  arrange(desc(percent_apo))

percent_apo_genus %>%
  filter(total_apo > 1) %>%
  group_by(genus) %>%
  summarize(
    n = mean(n),
    total_apo = mean(total_apo),
    percent_apo = mean(percent_apo),
    .groups = "drop"
  ) %>%
  mutate(
    genus =  paste0("*", genus, "*"),
    genus = fct_reorder(genus, percent_apo),
    label = glue::glue("{total_apo}/{n}")) %>%
  ggplot(aes(x = percent_apo, y = genus)) + 
  geom_col() +
  geom_text(aes(label = label), hjust = -.3) +
  scale_x_continuous(labels = percent, limits = c(0, .85)) +
  labs(x = "% apomictic taxa") +
  theme(
    axis.text.y = element_markdown(),
    axis.title.y = element_blank()
  )
```

**`r s_figure("apo-genus")`**. Rate of apomixis in genera of native, non-hybrid Japanese ferns with > 1 apomictic taxon. Numbers next to each bar show number of apomictic taxa out of total richness per genus.
